---
title: "Fig2 Optimum bivariate discretization for mutual information estimation"
output:
  github_document:
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The proposed information-maximizing discretization scheme is illustrated for a joint distribution defined as a Gumbel bivariate copula with parameter $\theta = 5$ and marginal distributions chosen as  Gaussian mixtures with three equiprobable peaks and respective means and variances,  $\mu_X = \{0,4,6\}$, $\sigma_X = \{1,2,0.7\}$ and $\mu_Y = \{-3,6,9\}$ , $\sigma_Y = \{2,0.5,0.5\}$.

```{r imports, message=FALSE}
library(MASS)
library(copula)
library(distr)
library(miic)
```

Defining marginals as multimodal distributions :
```{r marginals}
mixnorm1 <- UnivarMixingDistribution(Norm(0,1), Norm(4,2), Norm(6,0.7))
dmixnorm1 <- d(mixnorm1); qmixnorm1 <- q(mixnorm1); pmixnorm1 <- p(mixnorm1)


mixnorm2 <- UnivarMixingDistribution(Norm(-3,2), Norm(6,0.5), Norm(9,0.5))
dmixnorm2 <- d(mixnorm2); qmixnorm2 <- q(mixnorm2); pmixnorm2 <- p(mixnorm2)
```

Defining the 2d distribution with a [Gumbel Copula](https://en.wikipedia.org/wiki/Copula_(probability_theory)#Archimedean_copulas) :
```{r copula}
mv.NE <- mvdc(gumbelCopula(5), margins=c("mixnorm1","mixnorm2"), paramMargins=list(list(),list()))
```

Generating samples and running the dynamic discretization :
```{r plot, fig.width = 6, fig.height = 6}
N = 500
data = rMvdc(N, mv.NE)
print(discretizeMutual(data[,1], data[,2])$info)
```

