---
title: "FigS3 Interaction-dependent optimum discretization."
output:
  github_document:
    pandoc_args: --webtex
bibliography: 
  - "../../PLoS_SI_Supp.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Optimum bivariate partitions obtained from $N=1,000$ samples of two different joint distributions $P(X,Y)$ sharing the same sampling of $X$ taken from a uniform distribution on $[0, 0.3]$, but with different dependences for $Y$.
(A) $Y$ is defined as $\log(X) + \epsilon_1$, and (B) $Y$ is defined as $X^5 + \epsilon_2$, where $\epsilon_1$ and $\epsilon_2$ are Gaussian noise terms chosen so that the mutual informations of both examples are comparable, {\color{black}$I(X;Y)\simeq 0.75$}.

This example shows that the optimum partition for $X$ depends on its specific relation with $Y$ and needs to be discretized with finer partitions in (A) at low $X$ values for which $Y\simeq \log X$ varies the most and in (B) at higher $X$ values for $Y\simeq X^5$.


```{r, fig.height=6, fig.width=6}
library(miic)

N = 1000

X = runif(N, min = 0, max=0.3)
noise = c(rnorm(N, sd=0.4))
Y1 = log(X) + noise
miic::discretizeMutual(X, Y1)$infok
Y2 = X**5 + noise/1900
miic::discretizeMutual(X, Y2)$infok
```
